---
title: 'The Pronunciation of the Word "Māori"'
author: ""
date: "30/08/2023"
output:
  bookdown::html_document2:
    fig_caption: true
    toc: true
    toc_float: true
    theme: flatly
    collapsed: no
    df_print: paged
    code_folding: show
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
    background-color: #95A044;
}

pre {
  max-height: 300px;
  overflow-y: 300px;
}
</style>
```
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  fig.align='center', 
  warning = FALSE,
  message = FALSE
)
```
# Overview

This document presents the analysis steps and results for 'The Pronunciation of the Word "Māori"'.

The following sections cover:

1. loading data and libraries,
2. initial data examination,
3. the binary flap model,
4. the binary diphthong model, and
5. the diphthong ratio model.

***NB:*** we are not able to share the data for this project due to
research ethics commitments. We provide this markdown to indicate the data
processing steps which we have carried out and to provide.


# Load data and libraries

We load the required libraries.

```{r}
# Tidyverse contains plotting and data manipulation functions used throughout
# this markdown file.
library(tidyverse)
library(gt) # gt provides additional table output options.

# Linear mixed modelling
library(lme4)
library(effects)
library(emmeans)
```

We read in the data and create a binary variable indicating whether a token 
is a diphthong or not. All diphthong tokens have a 'ratio' of /a:/ to /o/, so
we can create a binary variable by indicating 'not diphthong' for every token
without a ratio.

```{r}
merged_data <- read_rds('merged_data.rds')

merged_data <- merged_data |> 
  # Store flap info as a factor for modelling.
  mutate(
    Flap = as.factor(Flap),
    generation = factor(
      merged_data$generation, levels=c("Hist", "TK", "Pres", "L1", "L2")
    ),
    Ratio = as.numeric(str_replace_all(Ratio, ",", "\\."))
  )

filtered_data <- merged_data |>
  # Remove Tūhoe speakers (for models with gender)
  filter(
    !generation == "TK"
  ) |> 
  # create diphthong variable. Ngāi Tūhoe speakers were not included in the 
  # diphthong models as gender was found to be significant.
  mutate(
    diphthong = as.factor(if_else(is.na(Ratio), 'n', 'y'))
  )
```

# Data exploration

What variables do we have?
```{r}
column_descriptions_initial <- c(
  "Search term used in LaBB-CAT",
  "Number of result",
  "Name of transcript",
  "Speaker name",
  "Speaker gender (binary)",
  "Language being spoken",
  "Speaker year of birth",
  "Speaker region",
  "Start time of transcript line (s)",
  "End time of transcript line (s)",
  "LaBB-CAT token identifier",
  "URL to access token in LaBB-CAT",
  "Previous segment",
  "Word (with whitespace removed)",
  "Following segment",
  "Word from LaBB-CAT text layer",
  "Onset of word (s)",
  "Offset of word (s)",
  "Word without punctuation",
  "Binary flap variable (y/n)",
  "Length of /a:/",
  "Length of /o/",
  "Ratio of /a:/ to /o/",
  "Comments concerning token",
  "MAONZE group identifier",
  "Speaker code without language identifier",
  "MAONZE generation category",
  "Binary diphthong variable (y/n)"
)

data_table <- tibble(
  `Variable Name` = names(filtered_data),
  Type = reframe(filtered_data, across(everything(), class)) |> as_vector(),
  Description = column_descriptions_initial
)

data_table <- data_table %>% 
  gt() %>% 
  tab_header(
    title = "Variables in the data set"
  )

data_table
```


We check for balanced data across our classes.

```{r}
table(merged_data$participant_gender, merged_data$participant_language)
```

```{r}
table(
  merged_data$participant_gender, 
  merged_data$participant_language, 
  merged_data$generation
)
```

Note the absence of female speakers in the Tūhoe generation. As noted above,
we exclude Tūhoe data from models with gender in the predictor variables.

What does the raw data look like?

```{r}
merged_data |> 
  group_by(CleanSpeaker, generation, participant_gender, participant_language) |> 
  summarise(
    total = n(),
    flap = sum(Flap == "y"),
    prop = flap/total
  ) |> 
  ggplot(
    aes(
      x = interaction(generation, participant_gender),
      y = prop,
      colour = participant_language,
    )
  ) +
  geom_boxplot() +
  geom_jitter() +
  labs(
    title = "Generation and gender by flap proportion",
  ) +
  theme_bw()
```

The results for `Hist.F` look extreme. We double checked these values and found
no reason to doubt them.

We can also consider the variation between individual speakers in flap
proportion.
```{r speaker-props, fig.cap = "Proportion flaps for speakers across languages. Thick lines are means for each category of speaker."}
speaker_data <- merged_data |> 
  group_by(CleanSpeaker, generation, participant_gender, participant_language) |> 
  summarise(
    total = n(),
    flap = sum(Flap == "y"),
    prop = flap/total
  )

mean_props <- speaker_data |> 
  ungroup() |> 
  group_by(generation, participant_gender, participant_language) |> 
  mutate(
    prop = mean(prop)
  )

speaker_data |> 
  ggplot(
    aes(
      x = participant_language,
      y = prop,
      group = CleanSpeaker,
      colour = generation,
      shape = participant_gender
    )
  ) +
  geom_line(alpha = 0.4) +
  geom_point(alpha = 0.4) +
  geom_line(
    data = mean_props,
    linewidth = 2
  ) +
  geom_point(
    data = mean_props,
    size = 5
  ) +
  labs(
    title = "Speaker flap proportions across languages",
    subtitle = "Speaker values and means for each category",
    colour = "Generation",
    shape = "Gender",
    x = "Language",
    y = "Proportion of flapped /r/"
  ) +
  theme_bw()
```

Figure \@ref(fig:speaker-props) already indicates some differences across 
categories.

What about for the diphthong variable?
```{r speaker-diphs, fig.cap = "Proportion dipthongs for speakers across languages. Thick lines are means for each category of speaker."}
speaker_data <- filtered_data |> 
  group_by(CleanSpeaker, generation, participant_gender, participant_language) |> 
  summarise(
    total = n(),
    flap = sum(diphthong == "y"),
    prop = flap/total
  )

mean_props <- speaker_data |> 
  ungroup() |> 
  group_by(generation, participant_gender, participant_language) |> 
  mutate(
    prop = mean(prop)
  )

speaker_data |> 
  ggplot(
    aes(
      x = participant_language,
      y = prop,
      group = CleanSpeaker,
      colour = generation,
      shape = participant_gender
    )
  ) +
  geom_line(alpha = 0.4) +
  geom_point(alpha = 0.4) +
  geom_line(
    data = mean_props,
    linewidth = 2
  ) +
  geom_point(
    data = mean_props,
    size = 5
  ) +
  labs(
    title = "Speaker diphthong proportions across languages",
    subtitle = "Speaker values and means for each category",
    colour = "Generation",
    shape = "Gender",
    x = "Language",
    y = "Proportion of diphthongs"
  ) +
  theme_bw()
```

In Figure \@ref(fig:speaker-diphs), Diphthong rates seem high across the board with the exception of two male 
present day kaumātua while speaking English. 

Finally, what does the ratio variable look like?
```{r all-ratio}
filtered_data |> 
  ggplot(
    aes(
      x = Ratio,
      colour = participant_language
    )
  ) +
  geom_freqpoly() +
  facet_grid(rows = vars(generation), cols = vars(participant_gender))
```

There are no obvious patterns in the distributions of dipthong ratio. All values
look sensible.

# Flap model

## Without Tūhoe speakers

For all models, we start with a maximal structure with three way interactions
for language, gender, and generation, along with random intercepts and slopes
on language for each speaker. If none of the coefficients for a predictor are
significant at the 0.05 level, we remove the predictor from subsequent models.

The following code fits the first model.
```{r}
Flapmodel0 <- glmer(
  # Dependent variable
  Flap ~ 
    # Independent variables / Main effects / predictors
    generation * participant_gender * participant_language + 
    # Random effects
    (1 + participant_language|CleanSpeaker), 
  # Family binomial because it's a 'yes or no' modelling problem.
  family = "binomial", 
  # Tell it the dataset.
  data = filtered_data, 
  # Some magic words to give the model additional time to 'converge'.
  glmerControl(optimizer = "nlminbwrap", optCtrl = list(maxfun = 1e7)),
  # Optimizer chosen using 'allFit' function
)
```

Let's have a look at this model. First the summary.

```{r}
summary(Flapmodel0)
```

The following description was written from one author to another. We leave it
in in case it is useful for any readers.

These values are a bit inscrutable by themselves. We will plot some effects in 
a moment to help understand them. But some general patterns are clear.
The first thing to look at is the `(Intercept)` value. This is the estimate for
the base level of the `generation` variable (i.e., `r levels(filtered_data$generation)[[1]]`).

The historical female speakers have a statistically significant (see the star)
estimate of `-2.32`. 'Significance' here means that the model is confident
that the true value is not `0`. The fact that the estimate is lower than 0 
(i.e., is negative), means the model estimates female
historical speakers, speaking English, to have a lower than 50/50 chance of
flapping.

The other values in the summary are interpreted with respect to the base level.
We add or subtract the values to the base level. e.g. to get the estimate for
historical male speakers speaking English, we add the value for
`participant_genderM` (`0.21409`) to the coefficient above (`-2.32`) to get
(`r -2.32 + 0.21409`) this value is still negative, so the model still takes
male historical speakers to have a lower than 50/50 chance of flapping while
speaking English. But note, the value for `participant_genderM` is not 
statistically significant (there's no star in the relevant row of the table). 
So the model is not confident that there is in reality any difference in the 
proportion of flapping between the genders in the historical speakers.

Before visualising, we continue to backfit. Note that none of the three way
interactions terms in the summary above are statistically significant. We fit
a new model without three way interactions.

```{r}
Flapmodel1 <- glmer(
  # Dependent variable
  Flap ~ 
    # Independent variables / Main effects
    generation + participant_language + participant_gender +  
    generation : participant_language +
    generation : participant_gender +
    participant_language : participant_gender + 
    # Random effects
    (1 + participant_language|CleanSpeaker), 
  # Family binomial because it's a 'yes or no' modelling problem.
  family = "binomial", 
  # Tell it the dataset.
  data = filtered_data, 
  # Some magic words to give the model additional time to 'congerge'.
  glmerControl(optimizer = "nlminbwrap", optCtrl = list(maxfun = 1e7))
)
```

```{r}
summary(Flapmodel1)
```

Now note that none of the terms with gender appear significant. We remove these
too.

```{r}
Flapmodel2 <- glmer(
  # Dependent variable
  Flap ~ 
    # Independent variables / Main effects
    generation * participant_language +
    # Random effects
    (1 + participant_language|CleanSpeaker), 
  # Family binomial because it's a 'yes or no' modelling problem.
  family = "binomial", 
  # Tell it the dataset.
  data = filtered_data, 
  # Some magic words to give the model additional time to 'congerge'.
  glmerControl(optimizer = "nlminbwrap", optCtrl = list(maxfun = 1e7))
)
```

```{r}
summary(Flapmodel2)
```
All terms are significant in this model. 

We now turn to plotting model
estimates.

```{r flap-model-out, fig.cap="Predicted flap proportions with 95% confidence intervals."}
mod_effects <- effect(
  term='generation:participant_language',
  mod=Flapmodel2
)

to_plot <- as_tibble(mod_effects)

to_plot |> 
  ggplot(
    aes(
      x = generation,
      y = fit, 
      colour = participant_language
    )
  ) +
  # add points
  # 'position_dodge' separates the two languages on the plot.
  # If you increase the width, the points get further apart.
  # If you change the value, change it in the code for the error bars
  # below as well.
  geom_point(position = position_dodge(width=0.5)) +
  # add errorbars
  geom_errorbar(
    aes(
      ymin = lower, 
      ymax = upper
    ),
    position = position_dodge(width=0.5),
    width = 0.4 # width of error bars.
  ) +
  labs(
    y = "Proportion of flapped /r/ (predicted)",
    x = "Generation",
    colour = "Language"
  ) +
  theme_bw() +
  scale_colour_grey(start = 0.6, end = 0)
```

Figure \@ref(fig:flap-model-out) provides a much clearer indication of what
is going on in this model. There is a large difference in flap proportion 
when speaking English and when speaking Māori. This difference seems reduced in
the present day kaumātua, and is not present at all in the contemporary L1 and
L2 speakers.

We generate p-values for these contrasts using the `emmeans` package.

```{r}
model.emm <- emmeans(Flapmodel2, specs = ~ generation * participant_language)
contrast(model.emm, "pairwise", simple = "participant_language")
```
These p-values match our visual judgement, where the historical and present
day generations have p-values less than 0.05 for the contrast between English
and Māori.

As a final check, to ensure we have the right model, we use the `anova` function.
```{r}
anova(Flapmodel0, Flapmodel1, Flapmodel2)
```
The results above indicate that our final model is the best of the models
considered. 

## With Tūhoe speakers

Since gender is not significant, we add the Tūhoe data back into the model.

```{r}
Flapmodel2Tūhoe <- glmer(
  # Dependent variable
  Flap ~ 
    # Independent variables / Main effects
    generation * participant_language +
    # Random effects
    (1 + participant_language|CleanSpeaker), 
  # Family binomial because it's a 'yes or no' modelling problem.
  family = "binomial", 
  # Tell it the dataset.
  data = merged_data, 
  # Some magic words to give the model additional time to 'congerge'.
  glmerControl(optimizer = "nlminbwrap", optCtrl = list(maxfun = 1e7))
)
```

```{r}
summary(Flapmodel2Tūhoe)
```
All terms again come out as significant. But is there a significant difference
for Tūhoe of language? We again plot and use `emmeans`.

```{r tuhoe-output, fig.cap = "Model predictions for flap variable with Tūhoe data added."}
mod_effects <- effect(
  term='generation:participant_language',
  mod=Flapmodel2Tūhoe
)

to_plot <- as_tibble(mod_effects)

figure_6 <- to_plot |> 
  mutate(
    participant_language = if_else(
      participant_language == "Maori", "Māori", participant_language
    )
  ) |> 
  ggplot(
    aes(
      x = generation,
      y = fit, 
      colour = participant_language
    )
  ) +
  # add points
  # 'position_dodge' separates the two languages on the plot.
  # If you increase the width, the points get further apart.
  # If you change the value, change it in the code for the error bars
  # below as well.
  geom_point(position = position_dodge(width=0.5)) +
  # add errorbars
  geom_errorbar(
    aes(
      ymin = lower, 
      ymax = upper
    ),
    position = position_dodge(width=0.5),
    width = 0.4 # width of error bars.
  ) +
  labs(
    y = "Proportion of flapped /r/ (predicted)",
    x = "Generation",
    colour = "Language"
  ) +
  theme_bw()  +
  scale_colour_grey(start = 0.6, end = 0)

# save for publication
ggsave(
  'figure_6.tiff',
  plot = figure_6,
  units = "in",
  dpi = 500,
  width = 7,
  height = 5
)

figure_6
```

Figure \@ref(fig:tuhoe-output) suggests that there is no significant effect
of language for the Tūhoe speakers.

We again use `emmeans`
```{r}
model.emm <- emmeans(Flapmodel2Tūhoe, specs = ~ generation * participant_language)
contrast(model.emm, "pairwise", by = "generation")
```
Our visual judgement of no significant effect of language for Tūhoe speakers is 
confirmed by the p-value above (0.1458).

Is there any evidence of difference in flapping between generations _within_
each language?
```{r}
contrast(model.emm, "consec", by = "participant_language", adjust = "bonferroni")
```
Tūhoe differ in their flap proportion while speaking te reo Māori.

# Binary diphthong analysis

We now apply the same approach to the binary diphthong variable. 

```{r}
Dmodel0 <- glmer(
  # Dependent variable
  diphthong ~ 
    # Independent variables / Main effects
    generation * participant_gender * participant_language + 
    # Random effects
    (1 + participant_language|CleanSpeaker), 
  # Family binomial because it's a 'yes or no' modelling problem.
  family = "binomial", 
  # Tell it the dataset.
  data = filtered_data, 
  # Some magic words to give the model additional time to 'converge'.
  glmerControl(optimizer = "nlminbwrap", optCtrl = list(maxfun = 1e7)),
  # Optimizer chosen using 'AllFit' - most don't fit as we're at the limits of 
  # what we can fit with these data.
)
```

Let's have a look at the summary.

```{r}
summary(Dmodel0)
```

No terms appear as significant, so we leave this variable behind.

# Diphthong ratio model

We now backfit using the diphthong ratio variable. This time we use a linear 
model, as the data is continuous.

```{r}
Rmodel0 <- lmer(
  # Dependent variable
  Ratio ~ 
    # Independent variables / Main effects
    generation * participant_gender * participant_language + 
    # Random effects
    (1 + participant_language|CleanSpeaker), 
  # Tell it the dataset.
  data = filtered_data, 
  # Some magic words to give the model additional time to 'converge'.
  ## lmerControl(optimizer = "nlminbwrap", optCtrl = list(maxfun = 1e7)),
  # Optimizer chosen using 'AllFit' - most don't fit as we're at the limits of 
  # what we can fit with these data.
)
```

```{r}
summary(Rmodel0)
```
These linear models do not give p-values. We use t-values great than 2 as a 
rough heuristic for keeping terms in our model. In this case, each term has
at least one 'significant' coefficient.

We check the residuals of the model to ensure that they are evenly distributed
around 0:
```{r}
plot(Rmodel0)
```
And check the qq plot:
```{r}
qqnorm(residuals(Rmodel0))
qqline(residuals(Rmodel0), datax = FALSE, distribution = qnorm,
       probs = c(0.25, 0.75), qtype = 7, col="red")
```

Both look OK!

We plot contrasts:

```{r}
mod_effects <- effect(
  term='generation:participant_gender:participant_language',
  mod=Rmodel0
)

to_plot <- as_tibble(mod_effects)

figure_7 <- to_plot |> 
  mutate(
    participant_language = if_else(
      participant_language == "Maori", "Māori", participant_language
    )
  ) |> 
  ggplot(
    aes(
      x = generation,
      y = fit, 
      colour = participant_language
    )
  ) +
  # add points
  geom_point(position = position_dodge(width=0.5)) +
  # add errorbars
  geom_errorbar(
    aes(
      ymin = lower, 
      ymax = upper
    ),
    position = position_dodge(width=0.5),
    width = 0.4 # width of error bars.
  ) +
  labs(
    y = "Ratio of diphthong (predicted)",
    x = "Generation",
    colour = "Language"
  ) +
  theme_bw()  +
  scale_colour_grey(start = 0.6, end = 0) +
  # split by gender
  facet_grid(rows = vars(participant_gender)) 

# save for publication
ggsave(
  'figure_7.tiff',
  plot = figure_7,
  units = "in",
  dpi = 500,
  width = 7,
  height = 5
)

figure_7
```

This is a complicated story!

```{r}
model.emm <- emmeans(Rmodel0, specs = ~ generation * participant_gender * participant_language)
contrast(model.emm, "consec", simple="each",  adjust="bonferroni")
```

In most cases with a significant language effect, there is an increase in the 
relative length of /a:/. The exception to this is the historical female 
speakers, who seem to reduce the relative length of /a:/ while speaking
te reo Māori. These results are discussed further in the paper. 